---
title: "Theme"
author: "AJ Thurston"
format:
  gfm:
    warning: FALSE
    message: FALSE
---

## Setup

```{r setup}
library(tidyverse)
library(ggplot2)
devtools::source_url("ajthurston.com/theme")
```

## Data
```{r data}
dat <- data.frame(values = as.factor(c(rep(0,6),rep(1,9))))
```

## Vertical Bar Plot

```{r plot}
ggplot(dat, aes(x = values)) +
  geom_bar(fill = "#336666", width = .5) +
  labs(x = "Values", y = "Frequency") +
  scale_y_continuous(limits = c(0,10), breaks = c(0:10), expand = c(0,0)) +
  theme_custom()
```

## Recoding Data

Horizontal bar plots are useful when the variable values are long strings of text.

The following recodes the dataset from numeric values 0 and 1 to strings.

```{r recode}
dat$values_name <- dat$values |>
  recode('0' = "This is a value of zero (0)\nin a long label format",
         '1' = "This is a value of one (1)\nin a long label format")
```

## Horizontal Bar Plot

Compare and contrast the vertical bar plot code above with the horizontal bar plot code below. Note that the data is now explictly plotted from the y-axis and frequencies are plotted on the x-axis.

```{r horizontal_bar}
ggplot(dat, aes(y = fct_rev(values_name))) +
  geom_bar(fill = "#336666", width = .5) +
  labs(x = "Values", y = "Frequency") +
  scale_x_continuous(limits = c(0,10), breaks = c(0:10), expand = c(0,0)) +
  theme_custom() +
  theme(axis.title.y = element_blank())
```